---
title: "LA3 One4ALL"
author: "Braden Griebel,Luke Fanning, Andrew Duffy, Ahyo Falick"
date: "9/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Adding data into cdc tibble
library(tidyverse)
source("http://www.openintro.org/stat/data/cdc.R")
cdc <- as_tibble(cdc)
```

```{r}
head(cdc)
```

# Team Question

**Question**: What effect does having a health plan have on general health, when controlling for smoking and gender?

**Reason**: We would like to discover whether or not health plans have a noticeable effect on people's health, as this could reveal whether efforts to increase the number of people with a health plan would actually be useful. Additionaly, by controlling for smoking and gender, it is possible to examine in what categories of people health plans make the most difference. Understanding this is important because it could reveal where efforts should be focused to increase the proportion of a specific group with healthplans in order to maximize the increase in health in society as a whole. This would also be useful in maximizing the improvement of average population health per money invested into this program, as obviously the groups that will benefit the most from health care plans will be the most cost effective when considering the overall impact on general health. This research should lead to greater average health of the population.  

```{r}
ggplot(data=cdc)+geom_bar(mapping=aes(x=hlthplan, fill=genhlth),position='fill')+facet_grid(gender~smoke100, labeller = label_both)+labs(x='Health Plan',y="Proportion",title="Health and Health Plans")
```

## Conclusion:
First and foremost, it is clear that across all the groups, those with a health plan have a higher proportion of people in the excellent and very good health categories, and smaller amounts in the fair and poor categories. This shows that having a health plan tends to increase the health category that people are in. Since this is true across all the groups that were controlled for, it shows that this is true regardless of gender and smoking habit. However, the actual amount of the increase in those categories varies by different amounts. The change doesn't vary much between males and females, but it is a significantly smaller increase in those who have smoked 100 cigarettes in their lifetime.It is also relatively obvious that regardless of whether or not a respondent has a health plan or is male or female, there are significantly more respondents in the poor health category among smokers versus non-smokers.

**Recommendation:** Due to the increase in health seen between those without healthplans and those that have healthplans, we recommend efforts be made to increase the amount of the population with health plans. Also, as the actual benefits of a healthplan vary between different groups, we recomend that more data is collected (or that the data in the BRFSS we didn't have access to be made more widely available) and analyzed to see which groups benefit most from healthplans. Then, those groups can be targeted by either campaigns to increase their desire for healthplans by displaying the potential benefits of having a health plan, or by making healthplans easier to access for those groups, most likely through reducing the cost to those in need who do not have the financial means necessary to pay for a health care plan.

# Braden Griebel Individual Question

```{r}
ggplot(data=cdc)+geom_boxplot(mapping=aes(x=gender,y=weight-wtdesire, fill=(smoke100==1)),position="dodge",alpha="0.7")+coord_cartesian(ylim=c(-10,50))+labs(x="Gender",y="Weight loss Desire", fill="Smoked 100 Cigarettes",title="Gender, Health and Weight loss Desire")+facet_wrap(~genhlth,nrow=2)
```


### Findings:
*In this boxplot it can be seen that for all categories, with the single excpetion of poor health males, those who have smoked more than 100 cigarettes want to lose at least as much weight as those who didn't smoke 100 cigarettes. Also, across all categories females want to loose more weight than the males. Finally, those in better health categories have a smaller difference between their current weight and their deesired weight.*

# Luke Fanning Individual Plot

```{r}
ggplot(data=cdc) + geom_point(mapping = aes(x=age, y= weight, color = age), position = "jitter") + facet_grid(hlthplan ~ ., labeller = "label_both") + theme(legend.position= "bottom") + labs(x="Age", y = "Weight", color = "", title = "Weight vs. Age, controlling for Health Care Plan")
```

### Findings:
*In this scatterplot, we are able to see a few harrowing discoveries. First, and perhaps more obvious, is that as age passes a certain limit, around 70 years or so, every respondent weighs under 250 pounds, regardless of if they have a health care plan or not, indicating just how much of an effect fitness and having a healthy lifestyle and weight have on lifespan. The second key takeaway from this graph also relates to lifespan: The number of people past 60 years of age in respondents without a health care plan is minute compared to the number of respondents with a health care plan, even in comparison to the smaller number of respondents witout a health care plan in general, and of those respondents, virtually all weigh between 100 to 200 pounds. While this could simply speak to the necessity of having a health care plan past a certain age, it more likely indicates that without a health care plan, a person must maintain a healthy weight if they wish to have a long lifespan whether or not they are unable or simply unwilling to get a health care plan.*


# Ahyo Falick Individual Question

```{r}
ggplot(data=cdc) + geom_smooth(mapping=aes(x=age,y=wtdesire,linetype=(exerany==1))) + labs(title = "Age vs. Weight Desire", caption = "Data source: The Behavioral Risk Factor Surveillance System (from a survey of 20,000 people)", linetype="Exercise", x="Age", y="Weight Desire")
```

### Findings:
*This graph represents the correlation between age and weight desire. It can be observed that people who exercise more regularly have a higher weight desire on average than those who don't with the exception of old age. There is a sharp decrease in weight desire among people over the age of 75 and an increase of people up to the age of 50. The data reaches a maximum at age 50, where both exercisers and non-exercisers desire a weight of about 157 pounds. The minimum weight desire occurs when people are 90+ years of age and they want to be, on average, 135 pounds.*



## Summary of Student Contributions:
I, Luke Fanning, used the geom point function to analyze the relationship between weight and age of respondents when controlling for health care plan. The data showed a clear positive relationship between maintaining a healthy weight and having a longer lifespan. I adjusted the color of the points as well as some of the label locations to make the graph easier to read, and used a facet grid model to display the health care plan more effectively alongside the given two variables. 

Ahyo Falick Contribution:
I used the geom smooth function to observe the correlation between age and weight desire among the 20,000 people surveyed. The linetype aesthetic was also used while splitting the group between those who exercised in the last month and those who did not. I added a caption detailing from where the data was acquired as well as labels for the axes and linetype to make the data easier to read. The data showed that people generally have an increasing desired weight up until the age of 50 where it decreases until their death. Additionally, people who exercise have a higher average desired weight than those who do not until around their 80's.


