---
title: "LA3 One4ALL"
author: "Braden Griebel,Luke Fanning, Andrew Duffy, Ayho Falick"
date: "9/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Adding data into cdc tibble
library(tidyverse)
source("http://www.openintro.org/stat/data/cdc.R")
cdc <- as_tibble(cdc)
```

```{r}
head(cdc)
```
# Braden Griebel Individual Question

```{r}
ggplot(data=cdc)+geom_boxplot(mapping=aes(x=gender,y=weight-wtdesire, color=(smoke100==1)))+coord_cartesian(ylim=c(-10,50))+labs(x="Gender",y="Weight loss Desire", color="Smoked 100 Cigarettes")+facet_wrap(~genhlth)
```


# Team Question

Question: What effect does having a health plan have on general health, when controlling for smoking and gender?
Reason: Whether or not health plans have a noticeble effect on peoples health could reveal whether efforts to increase the number of people with a health plan would actually be useful. This could effect 
```{r}
ggplot(data=cdc)+geom_bar(mapping=aes(x=hlthplan, fill=genhlth),position='fill')+facet_grid(gender~smoke100)+labs(x='Health Plan',y="Proportion",title="Health and Health Plans")
```

