---
title: "LA15 one4all"
output: html_document
---
Andrew Duffy, Braden Greibel, Luke Fanning, Ahyo Falick.

### This dataset comes from data.worldbank.org. 
It is a very large selection of statistics with a domain of almost every country, worldwide. It ranges from years 1995-2019.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(purrr)
library(modelr)

```

```{r message=F , warning=F}
perm_mean <- function(perms = 1000, all_values, n_A)
{
  ## Variables ##
  # perms: The number of permutations 
  # all_values (num): all data values
  # n_A (int): Size of group A
  ###############
  
  # Step 1:
  # Create vector of zeroes of length "perms" to store
  # permuted mean differnces
  permuted_mean_differences<-rep(0,perms)
  
  # Loop throught number of permutations
  for (i in c(1:perms))
  {
    # Step 2:
    # Randomly separate vector "values" into disjoint 
    # groups of size "nA" and "nB" respectively
    temp<-sample(all_values)
    group_A<-temp[1:n_A]
    group_B<-temp[(n_A+1):length(temp)]
    
    # Step 3:
    # Compute the sample means for the two groups from 
    # step 2
    mean_A<-mean(group_A,na.rm=T)
    mean_B<-mean(group_B,na.rm=T)
    
    
    # Step 4: 
    # Compute the difference in sample means, store the
    # value in the vector from step 1
    permuted_mean_differences[i]<-mean_A-mean_B
    
  }
  
  # Step 5:
  # Return the permuted mean differences vector
  return(permuted_mean_differences)
  
}
```


```{r message=F , warning=F}
perm_cor <- function(perms = 1000, x, y)
{
  ## Variables ##
  # perms: The number of permutations 
  # x: Vector of Variable 1 - for computing correlation
  # y: Vector of Variable 2 - for computing correlation
  ###############
  
  # Step 1:
  # Create vector of zeroes of length "perms" to store
  # permuted mean differnces
  correlations<-rep(0,perms)
  
  
  # Loop throught number of permutations
  for (i in c(1:perms))
  {
    # Step 2:
    # Randomly mix up the values in the vector "y"
    temp<-sample(y)
    
    # Step 3:
    # Compute the correlation between x and the randomly mixed
    # up y-vector. Store this value in the vector from step 1.
    correlations[i]<-cor(x,temp,use="pairwise.complete.obs")
    
  }
  
  # Step 4:
  # Return new updated vector, created in step 1
  return(correlations)
  
}

```



```{r}
df <- read_csv(file = "data1.csv", col_types=cols(CreationDate=col_datetime()))
info <- read_csv(file = "data2.csv")

colnames(df) <- c("Country", "Abbreviation", "Measuring", "Code", "1995", "1996", "1997", "1998", "1999", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019")

worlddata <- df %>% group_by(Country) %>% nest()
df<-df[,-29]

df<-df%>%gather(`1995`:`2019`,key=year,value=measurement)%>%select(Country,Measuring,year,measurement)%>%filter(Country!="Data from database: World Development Indicators")%>%
  filter(Country!="Last Updated: 10/28/2019")%>%filter(!is.na(Country))


countries<-unique(df$Country)
#countries

afghanistan<-df%>%filter(Country=="Afghanistan")%>%spread(key=Measuring,value=measurement)

cleaned<-afghanistan

countries<-countries[-1]
for(i in 1:length(countries)){
  temp<-df%>%filter(Country==countries[i])%>%spread(key=Measuring,value=measurement)
  cleaned<-rbind(cleaned,temp)
}


cleaned
```

# Team Question:

Does female HIV rate have a correlation with life expectancy and literacy rate? Expect the correlation to be inverse, where an increase in HIV rate leads to a decrease in life expectancy.

Null Hypothesis: There is no inverse correlation between HIV rate and test statistics Literacy and Life expectancy.

Alternate Hypothesis: There is an inverse correlation between HIV rate and test statistics life expectancy and literacy rate.

*note: measuring female stastic of HIV rate because male data is unavailable.



```{r}
tdata<-cleaned%>%select(Country,year,GDPpc=`GDP per capita (current US$)`,Literacy= `Literacy rate, adult total (% of people ages 15 and above)`,Life_exp=`Life expectancy at birth, total (years)`,
                        HIV=`Women's share of population ages 15+ living with HIV (%)`)

tdata<-tdata%>%mutate(year=as.double(year),GDPpc=as.double(GDPpc),Literacy=as.double(Literacy),Life_exp=as.double(Life_exp),HIV=as.double(HIV))

tdata

```

First, we should graph the data, by year. Then, we will perform a permutation correlation test with the variables we wish to find a correlation. 

We will start with graphing the yearly average of each statistic, worldwide, to get a look at the graphs and see if any graphs look correlated. We will add GDP per capita to get a decent understanding of the world's economic standing.

Two Statistics that jump out at me in terms of potential correlation is Life Expectancy and HIV rate. Note that HIV will have an inverse effect on Life expectancy until the life-saving medication is discovered for it. 

```{r, message = FALSE, warning = FALSE}
visualdata <- tdata %>% group_by(year)

visualdata <- visualdata[, -1]


ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = HIV), se = FALSE) + ggtitle("Worldwide rate of Women with HIV") + labs(x = "Year", y = "Female HIV rate")

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = Life_exp), se = FALSE) + ggtitle("Worldwide rate of Life Expectancy") + labs(x = "Year", y = "Life Expectancy")

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = GDPpc), se = FALSE) + ggtitle("Worldwide average of GDP per capita") + labs(x = "Year", y = "GDP per capita (US$)")

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = Literacy), se = FALSE) + ggtitle("Worldwide Literacy rate") + labs(x = "Year", y = "Literacy rate")


```

As we can see, the graph of hiv rate has a slight upward pointing "bump" in the trend line that is otherwise linear. The graph of life expectancy has a inversely related shift, pointing at a potential negative correlation before life-saving medication is created (in the mid 2000's).

Now that we have visualized the worldwide averages of our test statistics, lets take a look at a few different countries. Since HIV is most rampant in Africa, lets take a look at the same test statistics for South Africa (a country very devastated by HIV). Then, since worldwide prevelance of HIV is pretty consistent accros Europe and North America, lets take a look at a country in the EU, such as the UK.

```{r, message = FALSE, warning = FALSE}
sadata <- tdata %>% filter(Country == "South Africa") %>% group_by(year)

ggplot(data = sadata) + geom_smooth(mapping = aes(x = year, y = HIV), se = FALSE) + ggtitle("South Africa rate of Women with HIV") + labs(x = "Year", y = "Female HIV rate")

ggplot(data = sadata) + geom_smooth(mapping = aes(x = year, y = Life_exp), se = FALSE) + ggtitle("South Africa Life Expectancy") + labs(x = "Year", y = "Life Expectancy")

ggplot(data = sadata) + geom_smooth(mapping = aes(x = year, y = GDPpc), se = FALSE) + ggtitle("South Africa GDP per capita") + labs(x = "Year", y = "GDP per capita")

ggplot(data = sadata) + geom_smooth(mapping = aes(x = year, y = Literacy), se = FALSE) + ggtitle("South Africa Literacy rate") + labs(x = "Year", y = "Literacy rate")

```

We can interpret from South Africa that the graphs of HIV rate and Life Expectancy have an extremely large negative correlation up until around 2006.


```{r, message = FALSE, warning = FALSE}
ukdata <- tdata %>% filter(Country == "United Kingdom") %>% group_by(year)

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = HIV), se = FALSE) + ggtitle("Worldwide rate of Women with HIV") + labs(x = "Year", y = "Female HIV rate")

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = Life_exp), se = FALSE) + ggtitle("Worldwide rate of Life Expectancy") + labs(x = "Year", y = "Life Expectancy")

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = GDPpc), se = FALSE) + ggtitle("Worldwide average of GDP per capita") + labs(x = "Year", y = "GDP per capita")

ggplot(data = visualdata) + geom_smooth(mapping = aes(x = year, y = Literacy), se = FALSE) + ggtitle("Worldwide Literacy rate") + labs(x = "Year", y = "Literacy rate")

```


UK results are very similar to worldwide results, especially in terms of HIV rate and Life expectancy. Where there is a slight exponential increase in HIV rate, there is a slight exponential decrease in Life Expectancy. A small but not negligible negative correlation is visible from the graphs of these test statistics.


Now for the permutation test.

```{r}
lit<-tdata$Literacy
HIV<-tdata$HIV
lifeExp<-tdata$Life_exp
cor_lit_le<-cor(lit,lifeExp,use = "pairwise.complete.obs")
cor_HIV_le<-cor(HIV,lifeExp,use="pairwise.complete.obs")

pc_lit_le<-tibble(cor=perm_cor(perms=1000,lit,lifeExp))
pc_HIV_le<-tibble(cor=perm_cor(perms=1000,HIV,lifeExp))



plow_lit_le<-quantile(pc_lit_le$cor,probs=.025)[[1]]
phigh_lit_le<-quantile(pc_lit_le$cor,probs=.975)[[1]]
  
plow_HIV_le<-quantile(pc_HIV_le$cor,probs=.025)[[1]]
phigh_HIV_le<-quantile(pc_HIV_le$cor,probs=.975)[[1]]

ggplot(pc_lit_le,aes(cor,fill=(cor<plow_lit_le|cor>phigh_lit_le)))+geom_histogram(binwidth = .01)+geom_vline(xintercept = cor_lit_le)+
  labs(title="Literacy rate and Life Expectancy Permutation Correlation Test",
       x="Correlation Between Literacy rate and Life Expectancy",
       fill="Tails")


ggplot(pc_HIV_le,aes(cor,fill=(cor<plow_HIV_le|cor>phigh_HIV_le)))+geom_histogram(binwidth = .01)+geom_vline(xintercept = cor_HIV_le)+
  labs(title="Female HIV rate and Life Expectancy permutation Correlation Test",
       x="Correlation between Female HIV rate and Life Expectancy",
       fill="Tails")

percentile_lit_le<-nrow(filter(pc_lit_le,cor<cor_lit_le))/nrow(pc_lit_le)
percentile_HIV_le<-nrow(filter(pc_HIV_le,cor<cor_HIV_le))/nrow(pc_HIV_le)


ptable<-tibble("Percentile of:"=c("Literacy Rate and Life Expectancy Permutation Correlation Test","Female HIV Rate and Life Expectancy Permutation Correlation Test"),"Value:"=
                 c(percentile_lit_le,percentile_HIV_le))
ptable



```







## Andrew Duffy Individual Question:

### Subquestion: 

Does urban population growth have an effect on hiv rate in Africa? That is, is there a correlation between female HIV rate and urban population growth?

Null Hypothesis: There is no correlation between female HIV rate and urban population growth in Africa.

Alternate Hypothesis: There is a significant correlation between female HIV rate and urban population growth in Africa.



Start off with extracting a data frame from the large set of data we cleaned up. 

```{r}
urbanHIV <- cleaned %>% select(Country,year, HIV=`Women's share of population ages 15+ living with HIV (%)`, Urban = `Urban population growth (annual %)`)

urbanHIV <- urbanHIV %>% mutate(year=as.double(year), HIV=as.double(HIV), Urban = as.double(Urban))
```


First, lets graph findings, for both worldwide and for a country that is devastated by aids, as well as a country not devastated by aids.

```{r}
ggplot(data = urbanHIV) + geom_smooth(mapping = aes(x = year, y = Urban), se = FALSE) + ggtitle("Worldwide Urban Population growth") + labs(x = "Year", y = "Urban Population Growth, %")

#we have already seen the worldwide hiv rate graph, but here it is again for good measure

ggplot(data = urbanHIV) + geom_smooth(mapping = aes(x = year, y = HIV), se = FALSE) + ggtitle("Worldwide rate of Women with HIV") + labs(x = "Year", y = "Female HIV rate")

```

As you can see, on a global scale there seems to be no correlation whatsoever between urban population growth and female HIV rate.

Now, lets test France and Mongolia.

```{r}
mongolia <- urbanHIV %>% filter(Country == "Mongolia")

ggplot(data = mongolia) + geom_smooth(mapping = aes(x = year, y = Urban), se = FALSE) + ggtitle("Mongolia's Urban Population growth") + labs(x = "Year", y = "Urban Population Growth, %")

ggplot(data = mongolia) + geom_smooth(mapping = aes(x = year, y = HIV), se = FALSE) + ggtitle("Mongolia's rate of Women with HIV") + labs(x = "Year", y = "Female HIV rate")
```

Visually, one can see that the graphs seem to have a similar path up until 2005-2006. Then the graphs become reciprocal of each other.

```{r}
Fr <- urbanHIV %>% filter(Country == "France")

ggplot(data = Fr) + geom_smooth(mapping = aes(x = year, y = Urban), se = FALSE) + ggtitle("France Urban Population growth") + labs(x = "Year", y = "Urban Population Growth, %")


ggplot(data = Fr) + geom_smooth(mapping = aes(x = year, y = HIV), se = FALSE) + ggtitle("France's rate of Women with HIV") + labs(x = "Year", y = "Female HIV rate")
```

As you can see, the graphs are similar in function to those of Mongolia


Now we will perform the permutation correlation test to see if there is any correlation between the two. We will run this with a sample of a few African countries then compare to France, since France seemed to have a surprisingly similar trend line for urban population growth and HIV rate. I will do the test up to 2006 because that is where HIV seems to be stop being a very life-threatening disease due to modern medicine and political acts, such as President George Bush's push to move high quantities of new HIV medication into Africa in 2003.

```{r}
AFsample <- urbanHIV %>% filter(year <= "2006") %>% filter(Country == "South Africa" | Country == "Mongolia" | Country == "Central African Republic" | Country == "Ethiopia")

AFRresults <- perm_cor(perms = 1000, AFsample$Urban, AFsample$HIV)

FRresults <- perm_cor(perms = 1000, Fr$Urban, Fr$HIV)

AFRresults
FRresults
```

in order to compare these results, we must find a single number to compare them with. I'll use the mean of the 1000 results for each, to compare the two. 

```{r}
afrmean <- mean(AFRresults)
frmean <- mean(FRresults)
```

The mean of the African sample test is:

```{r}
afrmean
```

The mean of the France test is:

```{r}
frmean
```

These results show that there is actually a greater correlation between HIV rate and urban population growth in France than in the sample of African countries. This is very surprising, but even more surprising is the fact that there seems to be a negative correlation in the African sample test. 


Both of these correlation results are extremely close to 0, so the null hypothesis would not be rejected anyways, but the fact that the correlation for the African sample is negative means there is no way to reject the null hypothesis.

These findings mean that there is no correlation between urban population growth and female HIV rate in Africa.























## Braden Griebel Individial Section:


### Question:
What is the relationship between Literacy and Life expectancy when accounting for 



### Using Linear Models to Gain Insight

```{r}
bdata<-tdata%>%select(-HIV)

bmodel<-lm(Life_exp~GDPpc,data=bdata)
bdata<-add_residuals(bdata,model=bmodel,var="resid_le")
bdata

ggplot(bdata,aes(x=Literacy,y=Life_exp,color=log(GDPpc,base = 10)))+geom_point()+geom_smooth(se=F,color="red")+labs(title="Literacy, Life Expectancy and GDP per capita",
                                                                                                    x="Literacy Rate",
                                                                                                    y="Life Expectancy",
                                                                                                    color="Base 10 Log of GDP",
                                                                                                    caption="Log used to show order of magnitude since it spans many orders of magnitude")                                                                                                    

ggplot(bdata,aes(x=Literacy,y=resid_le))+geom_jitter()+geom_smooth(se=F,color="red")+labs(title="Literacy and Residuals of Life Expectancy",
                                                                              caption="Residuals Generated After removing effect of GDP per Capita",
                                                                              x="Literacy Rate",y="Residuals")

ggplot(bdata,aes(x=Literacy,y=resid_le))+geom_jitter()+geom_smooth(se=F,color="red")+xlim(85,100)+labs(title="Literacy and Residuals of Life Expectancy", subtitle = "Zoomed into 85-100% Literacy",
                                                                              caption="Residuals Generated After removing effect of GDP per Capita",
                                                                              x="Literacy Rate",y="Residuals")


bmodel2<-lm(Life_exp~GDPpc+Literacy,data=bdata)
print("Coefficients of Linear Model Including Per Capita GDP and Literacty Rate")
tibble(Measurement=c("Intercept","GDP per Capita", "Literacy Rate"),"Coefficient Value"= coef(bmodel2))
```


### Is there a difference in the literacy rates between the highest 50% of Life Expectancy, and the Lowest 50%?

```{r}
high_le<-bdata%>%filter(Life_exp>median(bdata$Life_exp,na.rm=T))
low_le<-bdata%>%filter(Life_exp<=median(bdata$Life_exp,na.rm=T))
c_mean_diff<-mean(high_le$Literacy,na.rm=T)-mean(low_le$Literacy,na.rm=T)


permuted_means<-tibble(lit_mean_diff=perm_mean(perms=1000,all_values = c(low_le$Literacy,high_le$Literacy),n_A=nrow(high_le)))

p2.5<-quantile(permuted_means$lit_mean_diff, probs=.025)[[1]]
p97.5<-quantile(permuted_means$lit_mean_diff, probs=.975)[[1]]

ggplot(data=permuted_means,aes(lit_mean_diff,fill=(lit_mean_diff<p2.5|lit_mean_diff>p97.5)))+geom_histogram()+geom_vline(xintercept = c_mean_diff)+
  labs(title="Permuted Difference of Means",fill="Tails",x="Difference of the Means of Literacy")

percentile<-nrow(filter(permuted_means,lit_mean_diff<c_mean_diff))/nrow(permuted_means)

print(paste("The Computed difference of Means is in the ",percentile*100,"th percentile",sep = ""))

```

### Is there a correlation between Literacy and the Calculated Residuals?
```{r}
resids<-bdata$resid_le
lit<-bdata$Literacy
calculated_correlation<-cor(lit,resids,use="pairwise.complete.obs")
permuted_correlations<-tibble(correlations=perm_cor(perms=1000,lit,resids))
p2.5<-quantile(permuted_correlations$correlations, probs=.025)[[1]]
p97.5<-quantile(permuted_correlations$correlations, probs=.975)[[1]]

ggplot(data=permuted_correlations,aes(correlations,fill=(correlations<p2.5|correlations>p97.5)))+geom_histogram()+geom_vline(xintercept = calculated_correlation)+
  labs(title="Permuted Correlations Between Literacy and Life Expectancy",x="Correlations",fill="Tails")

percentile<-nrow(filter(permuted_correlations,correlations<calculated_correlation))/nrow(permuted_correlations)

print(paste("The Computed Correlation is in the ",percentile*100,"th percentile",sep = ""))

```

## Luke Fanning Individual Section

### Individual Question: 
```{r}
(ldata <- tdata %>% filter(year == 2018) %>% filter(!is.na(Life_exp))%>% arrange(desc(Life_exp)) %>% group_by(Life_exp, Country) %>% head(10)) %>% select("Country", "Life_exp")
```

The tibble above shows the top ten countries in terms of Life expectancy in 2018, the most recent year in which life expectancy data is available in this dataset. We can also view each of these countries life expectancies over time, as seen below.
```{r}
l2data <- tdata %>% filter(Country %in% c("Hong Kong SAR, China", "Japan", "Macao SAR, China", "Switzerland", "Spain", "Italy", "Singapore", "Channel Islands", "Luxemborg", "Korea, Rep.")) %>% filter(!is.na(Life_exp)) %>% group_by(Life_exp, Country)

ggplot(l2data, aes(x = year, y = Life_exp, color = Country)) + geom_point() + labs(x = "Year", y = "Life Expectancy, in years", title = "Life Expectancy over time for top ten countries in 2018")
```

From the above graph, we see that the two countries with the most dramatic change in life expectancy over the time period included in the dataset are The Republic of Korea, more commonly referred to as South Korea, and Singapore. Obviously, it is fantastic that these countries have made such great improvements in their life expectancies over the last 23 years, but is this indicative of a greater trend, or are these countries outliers in terms of their improvements? 
In other words, is the change in life expectancy in these countries between 1995 and 2018(the scope of the data) expected when compared to the average change in life expectancy for all countries in this time period, or are these changes statistically significant? 
This question is important primarily due to the implications behind these countries being statistically significant in their changes, as if they are, it is important that we further explore exactly what policies and societal changes these countries have made in order to improve their life expectancy so dramatically up to world leading levels. These changes could then be implemented in other countries around the globe in order to allow for similar increases and lead to a longer living populace overall. 

### Analysis
```{r, include = F}
ldata1995 <- tdata %>% filter(year %in% c(1995))%>% filter(!is.na(Life_exp)) %>% select("Country", "Life_exp")
ldata2018 <- tdata %>% filter(year %in% c(2018))%>% filter(!is.na(Life_exp)) %>% select("Country", "Life_exp")

(l3data <- full_join(ldata1995, ldata2018, by = "Country") %>% mutate(Life_exp_difference = Life_exp.y - Life_exp.x)) %>% select("Country", "Life_exp_difference")

```
```{r, include = F}
real_diff_Singapore_Korea <- l3data %>% filter(Country %in% c("Singapore", "Korea, Rep."))
diff_singapore <- real_diff_Singapore_Korea %>% filter(Country == "Singapore") %>% pull(Life_exp_difference)
diff_S_Korea <- real_diff_Singapore_Korea %>% filter(Country == "Korea, Rep.") %>% pull(Life_exp_difference)
```

```{r}
ggplot(l3data, aes(x = Life_exp_difference)) + geom_histogram(binwidth = 1)+ geom_vline(xintercept = diff_S_Korea) + geom_vline(xintercept = diff_singapore) + labs(x = "Difference in Life expectancy in years", y = "Count", title = "Change in Life Expectancy between 1995 and 2018")
```

Clearly, from the above graph, it can be seen that these two differences are not overwhelmingly different from a lot of the changes in life expectancy around the world for the given time period. However, it can be noted that these changes both occured in Asian countries, and in fact, most of the highest life expectancy countries are in Asia, which then leads us to a second question worth evaluating: Is the difference in mean life expectancy change from 1995 to 2018 between Asian countries and the rest of the world statistically significant? 
Again. this question is important to evaluate, as it can be indicative of a continent making correct societal and lifestyle choices compared to the rest of the world. 
In this test, we will consider the null hypothesis to be that there is no difference in mean change in life expectancy from 1995 to 2018 between Asian countries and the rest of the world, and the alternative hypothesis to be that there is.

```{r, eval = F, include = F}

perm_mean <- function(perms = 1000, all_values, n_A)
{
  permuted_mean_differences <- numeric(perms)
  for (i in c(1:perms))
  {
    rand_vector <- sample(all_values) #shuffles all values as a vector
    group_A <- rand_vector[1:n_A]
    group_B <- rand_vector[(n_A+1):length(rand_vector)]
    mean_A <- mean(group_A)
    mean_B <- mean(group_B)
    difference <- mean_A - mean_B
    permuted_mean_differences[i] <- difference
  }
  return(permuted_mean_differences)
}
```

```{r}
l4data <- l3data %>% filter(!is.na(Life_exp_difference))
l3data_Asia <- l3data %>% filter(Country %in% c("China", "India", "Indonesia", "Pakistan", "Bangladesh", "Japan", "Philippines", "Vietnam", "Turkey", "Iran, Islamic Rep.", "Thailand", "Myanmar", "Korea, Rep.", "Iraq", "Afghanistan", "Saudi Arabia", "Uzbekistan", "Malaysia", "Yemen, Rep.", "Nepal", "Sri Lanka", "Korea, Dem. People’s Rep.", "Kazakhstan", "Syrian Arab Republic", "Cambodia", "Jordan", "Azerbaijan", "United Arab Emirates", "Tajikistan", "Israel", "Lao PDR", "Lebanon", "Kyrgyz Republic", "Turkmenistan", "Singapore", "Oman", "Kuwait", "Georgia", "Mongolia", "Armenia", "Qatar", "Bahrain"))
l3data_rest <- l3data %>% filter(!Country %in% c("China", "India", "Indonesia", "Pakistan", "Bangladesh", "Japan", "Philippines", "Vietnam", "Turkey", "Iran, Islamic Rep.", "Thailand", "Myanmar", "Korea, Rep.", "Iraq", "Afghanistan", "Saudi Arabia", "Uzbekistan", "Malaysia", "Yemen, Rep.", "Nepal", "Sri Lanka", "Korea, Dem. People’s Rep.", "Kazakhstan", "Syrian Arab Republic", "Cambodia", "Jordan", "Azerbaijan", "United Arab Emirates", "Tajikistan", "Israel", "Lao PDR", "Lebanon", "Kyrgyz Republic", "Turkmenistan", "Singapore", "Oman", "Kuwait", "Georgia", "Mongolia", "Armenia", "Qatar", "Bahrain"))
real_mean_diff <- mean(l3data_Asia$Life_exp_difference) - mean(l3data_rest$Life_exp_difference)
permuted_means <- perm_mean(1000, l4data$Life_exp_difference, 100)
percentile_Asia_vs._World <- length(permuted_means[permuted_means < real_mean_diff])/1000

tibble(val = permuted_means) %>% ggplot(., aes(val)) + geom_histogram(binwidth = 0.1) + geom_vline(xintercept = real_mean_diff) + labs(x = "Mean Change in life expectancy", y = "Count", title = "Mean Changes in life expectancy difference from 1995 to 2018 around the world")

```

When compared to the rest of the world, the mean difference in life expectancy change from 1995 to 2018 in Asia is actually -0.30, putting it in the 32nd percentile of all life expectancy changes. Therefore, the null hypothesis can be accepted. Clearly, while life expectancies in many Asian countries are higher, on average, there is no statistical evidence to suggest that they have increased more in the past 23 years compared to the rest of the world. This then suggests that while these countries are doing some things right in terms of societal choices and lifestyle changes, these choices and changes are not significant in affecting life expectancy compared to the choices and changes made across the rest of the world. While this may be discouraging in the sense that there is no one true answer to raising life expectancy, this is actually encouraging, as it suggests that the world overall is making the correct choices to increasing life expectancy and that we just need to be patient and continue to work on technology and improvement in our spheres of life. The data simply suggests that there is not one true answer, but rather a large number of answers for increasing life expectancy depending on the culture and society of the country currently working to fix that problem or improve their already high expectancy. 

## Reflections:
### Team Reflection:

In the beginning of the semester our team goal was to "To learn techniques to explore and understand data, as well as being able to present that data in an clear, attractive and intriguing way".

We feel that we have met that goal and feel comfortbale representing large amounts of data visually and with certain statistic tests to get desired results.












how have your six-month or 5-year goals changed? What did you learn/accomplish in this course? If you could give yourself advice at the beginning of the semester, what would you tell yourself to keep doing, stop doing, and start doing?
### Braden Reflection:
At the start of the semester my six month goal was "to be in grad school, I don’t know specifically for what, but hopeully something related to statistics." and my five year post graduation goal was "I hope to be working on epidemiology for the CDC".

###Andrew Reflection:
At the beginning of the semester my goals were to work in Europe after graduation to see what it is like to live in another country besides America. My long term goals were to have a comfortable job where I have the financial freedom to enjoy my hobbies, and to own my own consulting company in the very long term. I wouldn't say that my goals have changed, but there are some more things I would like to see out of my life. Something I would add is use my potential knowledge of statistics and data to better the world around me the best way that I can. I learned a lot in this course, an extremely valuable base knowledge of how to visually represent data and perform different tests on large groups of data via computing. If I could give myself any advice it would be to complete the labs earlier!

### Luke's Reflection:
Obviously, my goals have changed a lot from the start of the semester to now, as I have once again changed my major. While I did really want to be in a sports-related field working with statistics and data, I realize now that sports and sports statistics are more of a hobby for me rather than a career interest. I have chnaged my life around a lot and am continuing to explore my own passions and career aspirations, but I think college is the time to do this, and through my self-exploration and pursuit of knowledge, I will come out a better and more well-rounded person. Therefore, while my 6 month and 5 year goals are more up in the air now than they were at the beginning of the semester, I feel that leaving these more open to exploration is a better path. Therefore, my new 5 year goal is to simply be a more learned, knowledgeable, and well-rounded person than I am now. Overall, I feel like I have learnt a lot from this course with reagards to the R programming language, and I feel this knowledge will be incredibly helpful for me moving forward. If I could give any advice to myself at the start of the semester, it would be to take detailed notes on each and every chapter and exercise in the book, as these will come to be incredibly helpful when working through each and every lab.


