---
title: "LA12"
author: "Braden Griebel, Andrew Duffy, Ahyo Falick, Luke Fanning"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(lubridate)
library(gapminder)
```


```{r}
perm_mean <- function(perms = 1000, all_values, n_A)
{
  ## Variables ##
  # perms: The number of permutations 
  # all_values (num): all data values
  # n_A (int): Size of group A
  ###############
  
  # Step 1:
  # Create vector of zeroes of length "perms" to store
  # permuted mean differnces
  permuted_mean_differences<-rep(0,perms)
  
  # Loop throught number of permutations
  for (i in c(1:perms))
  {
    # Step 2:
    # Randomly separate vector "values" into disjoint 
    # groups of size "nA" and "nB" respectively
    temp<-sample(all_values)
    group_A<-temp[1:n_A]
    group_B<-temp[(n_A+1):length(temp)]
    
    # Step 3:
    # Compute the sample means for the two groups from 
    # step 2
    mean_A<-mean(group_A)
    mean_B<-mean(group_B)
    
    
    # Step 4: 
    # Compute the difference in sample means, store the
    # value in the vector from step 1
    permuted_mean_differences[i]<-mean_A-mean_B
    
  }
  
  # Step 5:
  # Return the permuted mean differences vector
  return(permuted_mean_differences)
  
}
```


```{r}
perm_cor <- function(perms = 1000, x, y)
{
  ## Variables ##
  # perms: The number of permutations 
  # x: Vector of Variable 1 - for computing correlation
  # y: Vector of Variable 2 - for computing correlation
  ###############
  
  # Step 1:
  # Create vector of zeroes of length "perms" to store
  # permuted mean differnces
  correlations<-rep(0,perms)
  
  # Loop throught number of permutations
  for (i in c(1:perms))
  {
    # Step 2:
    # Randomly mix up the values in the vector "y"
    temp<-sample(y)
    
    # Step 3:
    # Compute the correlation between x and the randomly mixed
    # up y-vector. Store this value in the vector from step 1.
    correlations[i]<-cor(x,temp)
    
  }
  
  # Step 4:
  # Return new updated vector, created in step 1
  return(correlations)
  
}
gapminder
```

```{r}
levels(gapminder$continent)
unique(gapminder$year)
```





```{r}
#gap_data<-read_delim(file="gapminder.csv",delim=",",col_types = cols(
#  country=col_factor(),
#  continent=col_factor(),
#  year=col_integer(),
#  lifeExp=col_double(),
#  pop=col_integer(),
#  gdpPercap=col_double()
#  
#))
gap_data<-gapminder
```


# Data Description:
The gapminder dataset has Life expectancy, population, and GDP per capita by year and country. The continent of the country is also recorded. The years range from 1952 to 2007.  

# Question 1: 
Is there a difference in the correlation between GDP and Life Expectancy between Europe and Asia?   
$H_0$:Cor(GDP per capita, Life expectancy) in Asia $-$ Cor(GDP per capita, Life expectancy) in Asia $=$ 0                
$H_1$:Cor(GDP per capita, Life expectancy) in Asia $-$ Cor(GDP per capita, Life expectancy) in Asia $\neq$ 0                
We will reject the null, and accept the alternative hypothesis and reject the null if and only if percentile of the difference in means is under $2.5\%$ or above $97.5\%$    

## Why this Question is Interesting:
There is a correlation between life expectancy and per capita GDP (can be shown with a permutation test), but is this correlation as strong in different regions? This is interesting because it can allow us to see what aspects can affect this correlation. This is important because it can allow for an understanding of how life expectancy is affected by a variety of dufferent variables. This in turn would allow for programs to reduce the disparity in life exxpectancy between the more and less wealthy countries (by increasing the life expenctancy of the less wealthy countries, not decreasing the life expectancy of the wealthier countries although that would be easier). 



```{r}
set.seed(25)
gap_data2<-gap_data%>%group_by(continent,country)%>%summarize(correlation=cor(gdpPercap,lifeExp))%>%filter(continent=="Asia"|continent=="Europe")

gap_dataAsia<-gap_data2%>%filter(continent=="Asia")
gap_dataEurope<-gap_data2%>%filter(continent=="Europe")
cor_diff<-mean(gap_dataEurope$correlation)-mean(gap_dataAsia$correlation)

permutation_diff<-perm_mean(perms=10000,all_values=gap_data2$correlation,n_A=length(gap_dataEurope$correlation))
permutaion_diff<-tibble("correlation"=permutation_diff)
ggplot(data=permutaion_diff)+geom_histogram(aes(x=correlation,fill=(correlation<quantile(correlation,probs = c(.025))[[1]]|correlation>quantile(correlation,probs = c(.975))[[1]])))+
  geom_vline(xintercept=cor_diff,color="red")+
  labs(title="Permutated Means Histogram",fill="Tails",x="Mean Difference")
percentile=sum(permutaion_diff<cor_diff)/nrow(permutaion_diff)*100
stats<-tribble(
  ~Statistic,              ~Value,
  #--|--|----
  "Mean", mean(permutaion_diff$correlation),
  "Median", median(permutaion_diff$correlation),
  "Max", max(permutaion_diff$correlation),
  "Min", min(permutaion_diff$correlation),
  "2.5 Percentile", quantile(permutaion_diff$correlation,probs=c(.025))[[1]],
  "97.5 Percentile", quantile(permutaion_diff$correlation,probs=c(.975))[[1]],
  "Calculated Mean Difference",cor_diff,
  "Percentile",percentile
)
stats
```



# Question 2:
Is there a correlation between population and GDP per capita for the African continent?                 
$H_0:\rho=0$                   
$H_1:\rho\neq0$           
We will reject the null, and accept the alternative hypothesis and reject the null if and only if percentile of the difference in means is under $2.5\%$ or above $97.5\%$                       



```{r}
i="Africa"
gap_data3<-gap_data%>%filter(continent==i)
thing1<-gap_data3$pop
thing2<-gap_data3$gdpPercap
correlations<-perm_cor(perms=10000,thing1,thing2)
correlations<-tibble("correlation"=correlations)
calculated_cor<-cor(thing1,thing2)
ggplot(data=correlations,aes(x=correlation,fill=(correlation<quantile(correlation,probs = c(.025))[[1]]|correlation>quantile(correlation,probs = c(.975))[[1]])))+geom_histogram()+
  geom_vline(xintercept=calculated_cor,color="red")+
  labs(fill="tails")
percentile<-length(correlations$correlation[correlations$correlation<calculated_cor])/nrow(correlations)
stats<-tribble(
  ~Statistic,              ~Value,
  #--|--|----
  "Mean", mean(correlations$correlation),
  "Median", median(correlations$correlation),
  "Max", max(correlations$correlation),
  "Min", min(correlations$correlation),
  "2.5 Percentile", quantile(correlations$correlation,probs=c(.025))[[1]],
  "97.5 Percentile", quantile(correlations$correlation,probs=c(.975))[[1]],
  "Calculated Correlation",calculated_cor,
  "Percentile",percentile
)
stats

```



# Individual Contributions:
I, Braden Griebel, implemented the perm_mean function and used it to run a permutation test on the correlation between Life expectancy and GDP per capita differed between Europe and Asia. I added colour to the graph to show where its tails were, in order to demonstrate whether the difference in mean correlation between Europe and Asia was significant. I also added a vertical line to show where the calculated difference in means lay.  



