---
title: "Lab 5"
date: "September 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overall Description

For this lab, you will perform exploratory data analysis on a dataset describing dairy production at the state level, dating back to 1970. The raw data was orignally collected and made public by the United States Department of Agriculture (USDA). Feel free to check it out [here](https://www.ers.usda.gov/data-products/dairy-data/). The data that we will be working with was cleaned and processed by contributers to the [tidy tuesday](https://github.com/rfordatascience/tidytuesday) github repository.

## Goals

- Identify overall trends in the milk production dataset.
- Hypothesize possible explanations for these trends and identify additional data sources that cound be helpful.

In order to get the data yourself, you will need to clone the repository using GitKraken or Github. Once you clone this repository, the dataset that you need to load is present at the following file location

- tidytuesday/data/2019/2019-01-29/state_milk_production.csv

Finally, let’s load the tidyverse library and view the milk production datset. Make sure that your script or markdown file is in the same file location as the data.

```{r, warning=FALSE, message=FALSE}
# Load tidyverse
library(tidyverse)

# Load and view the data
milk <- read.csv('state_milk_production.csv')
head(milk)
```

It appears that we have four columns of data, corresponding to region, state, year, and milk_produced (in pounds).One thing to notice is that the ‘milk_produced’ column has high magnitude. Let’s create a new column to indicate millions of pounds of milk. Our figures and tables will look nicer if we work with millions of pounds, rather than pounds.

```{r}
# Add a column for millions of milk pounds
milk <- milk %>%
  mutate(milk_million = milk_produced/1000000)
```


Now we have a column indicating millions of pounds of milk. Since we have taken care of that, let’s check out the initial year of data given, 1970. We can extract the data for that particular year with the following functions.

```{r}
# Extract the data for 1970 only
milk1970 <- milk %>%
  filter(year == 1970)
```


# Univariate Distribution

One usefule tool for exploratory data analysis is to visualize the distribution of your data. For continuous data, we can get an estimate for the univariate distribution of a variable by using geom_histogram and geom_density.

For the 1970s milk data, let’s plot a histogram for millions of pounds of milk used. Make sure you choose an appropriate number of bins to make the plot look good.

```{r}
ggplot(data = milk1970, aes(x = milk_million)) +
  geom_histogram(bins = 15) + 
  ggtitle('Histogram of milk produced in 1970 by state')
```

It appears that most states in 1970 produced about 2 billion pounds of milk. One state produced over 15 billion pounds of milk! Can you identify this state?

```{r}
milk1970 %>% filter(milk_million>15000) %>% print()
```
*The state that produced more than 15 billion pounds of milk was Wisconsin*

Another similar kind of plot is a probability density plot. Here is what a density plot looks like. With a density plot, we estimate a probability density function for the data that is presented

```{r}
ggplot(data = milk1970, aes(x = milk_million)) +
  geom_density() + 
  ggtitle('Distribution estimate of milk produced in 1970 by state')
```


# Comparing Univariate Distributions

Another kind of plot that is useful for comparing univariate distributions across different groups are boxplots. For example, we can filter the data for three years, say 1970, 1990, and 2010, to visualize how the distribution of milk procutions changes across those years.


```{r}
# Extract data for the years 1970, 1990, and 2010
boxMilk <- milk %>%
  filter(year %in% c('1970','1990','2010'))
```

Now that we have filtered our data, let’s create a boxplot

```{r}
ggplot(data = boxMilk, aes(x = as.factor(year), y = milk_million, fill = as.factor(year))) +
  geom_boxplot() + 
  ggtitle('Pounds of milk produced at the state-level') + 
  scale_fill_discrete(name = 'Year') + 
  xlab('Year') +
  ylab('Milk Produced (Millions lb)')
```

There are “outliers” present in the data that is disrupting the scale of the plot. Try to remedy this when you are creating your own plots. To interpret the above plot, it appears that the variation in the state-level milk production increases from 1970 to 2010. This is evident from the expanded inter-quartile range (colored area) as you cross the graph from left to right. We also see more substantial “outliers” in 2010, than we do in 1970. It appears the the median milk production stays approximately the same over the years. We can veryify this with the following computation.

```{r}
# Extract data for the years 1970, 1990, and 2010
median_year <- boxMilk %>%
  group_by(year) %>%
  summarise(Med = median(milk_million)) %>%
  as_tibble()
median_year
```

Other plots similar to a boxplot are violin plots and dot plots. You can try out geom_dotplot and related geom_violin. Generally, data scientists recommend a dotplot when there are not many data points to plot, and a violin plot when there are too many data points to plot.

# Bivariate Distributions

We haven’t even checked out the coolest part about this data: it’s time series data. We can plot the milk production over time to get an understanding of how consumption changed at the state level. Let’s plot the milk consumption over time for the states: New York, Wyoming, Maryland, Maine, and Florida. While we are at it, we only really need the columns state, year, and milk_million. We will only select these columns

```{r}
milksub <- milk %>%
  filter(state %in% c('New York','Wyoming','Maryland','Maine','Florida')) %>%
  select(state, year, milk_million)
```

Now finally, let’s plot the time series data for these 5 states.

```{r}
ggplot(data = milksub, aes(x = year, y = milk_million, color = state)) +
  geom_point() + 
  ggtitle('Pounds of milk over time by state') +
  xlab('Year') +
  ylab('Milk Produced (Million lb)')
```

What does this tell you about the milk production from 1970 to now?

*For most of the states in the set we are examining, milk production has decreased. However, for Florida and New York it has it increased. New Yorks rate of increase was significcantly larger than for Florida. Also, across all the years of the data, the ranking of milk production by state hasn't changed, going New York, Florida, Maryland, Maine, and finally Wyoming. New York also produced a lot more than any of the other states. *

Other useful plots for visualizing bivariate relationships are geom_line, geom_jitter, geom_bin2d, geom_hex, and others.

# Summarising Statistics

We did some summarising in a previous code-chunk, however we will formally cover it here. We will be using the function group_by and summarise to report correlations between year and milk production for our 5 states.

```{r}
cor_sum <- milksub %>%
  group_by(state) %>%
  summarise(Correlation = cor(year,milk_million)) %>%
  as_tibble()
cor_sum
```

The Pearson correlation give us a measurement on the strength of the linear relationship between two variables. For example, the correlation between year and milk produced for New York is .9, indicating that there is a strong linear relationship here. While for Wyoming, we obtain a correlation of -.25, indicating a weak linear realtionship.

**Note that correlations have nothing to do with the of the linear relationship between two variables.** For example, the correlation between year and milk produced for New York and Maryland are .9 and -.9 respectively. However, the size of this relationship between the two states is clearly different, as evident from the last graph.

Other useful summary statistics are:

- Mean
- Median
- Variance
- Standard Deviation
- Max
- Min

# Assignment Instructions

*Due: 5:00PM Sunday, Sept 29, 2019 on OSF*

Individually, perform three tasks. For the first, display the distribution of the yearly milk production (in millions of pounds) for your birth-year. For the second, summarise the data for your birth year. Report the average milk produced by each state, the median milk produced be each state, the state that produced the most milk, and the state that produced the least amount of milk.

As a team, select all of the states in which at least one of your group members has lived in. If there are less than 5, choose additional states so that you select 5 of them. For these states, plot a yearly time series for the milk production (in millions of pounds). **On this plot, display a line showing the average milk produced each year for all states.** Display this plot on the front page and comment on what information the plot conveys.

Then using the whole dataset, report the following on the front page.

- The year when the most milk was produced in the United States.
- The year when the least milk was produced in the United States.
- In 2017, report the 5 states that produced the most milk. (Hint: Use arrange)
- In 2017, report the 5 states that produced the least milk. (Hint: Use arrange)

After everyone is finished, as a team, write a few sentences in the team report in R Markdown detailing what you found to be the most helpful in learning about this data. Include all individual work in this document as well.

The final task is to reflect on the ethics of this data analysis. Brainstorm with your team on who could benefit most from this analysis? Who could suffer? Write a few sentences explaining who could benefit and suffer from this analysis and why.

# Individual Sections

## Braden Griebel


```{r}
milk %>% mutate(milk_million=milk_produced/1000000) %>% filter(year==1998) %>%   arrange(desc(milk_million))%>%
  ggplot(aes(x=reorder(state,milk_million,FUN=identity),y=milk_million))+geom_col(width=1,color="white")+coord_flip(ylim=c(1100,27000))+labs(y="Milk Produced (Millions of Pounds)",x="State", title="Milk Production by State in 1998")+ theme( axis.text.y = element_text(size=5))
```


### Summary of Milk Production for My Birth Year (1998)


```{r}
milk%>% filter(year==1998) %>% mutate(milk_million=milk_produced/1000000)%>%group_by(state)%>%summarize(average=mean(milk_million),median=median(milk_million))%>%as_tibble()%>%print()
```

```{r}
milk_ordered<-milk%>% filter(year==1998) %>% mutate(milk_million=milk_produced/1000000)%>%group_by(state)%>%arrange(milk_million)
```

### State that Produced the Least milk

```{r}
milk_ordered[1,c(2,5)]
```

### State that Produced the most milk
```{r}
milk_ordered[length(milk_ordered$state),c(2,5)]
```

## Ahyo Falick

```{r}
milk %>% mutate(mm=(milk_produced/1000000)) %>% filter(year==1999) %>% ggplot(aes(x = mm, color = state)) +
  geom_histogram(bins = 20) + theme(legend.position = "none") +
  labs(title = "Histogram of Milk Produced in 1999 by State", y = "Count", x = "Milk Produced (Millions)")
```

### Summary of Average Milk Production by State in 1999

```{r}
milk %>% mutate(mm = milk_produced/1000000) %>% filter(year == 1999) %>% group_by(state) %>% summarize(average=mean(mm),median=median(mm))%>%as_tibble()%>%print()
```

### State That Produced the Most Amount of Milk

```{r}
milk %>% filter(year == 1999) %>% arrange(desc(milk_million)) %>% filter(state == "California")

```

### State That Produced the Least Amount of Milk


```{r}
milk %>% filter(year == 1999) %>% arrange(desc(milk_million)) %>% filter(state == "Alaska")

```

## Luke Fanning

### Distribution of milk production in 1997

```{r}
milk %>% filter(year == 1997) %>%  mutate(milk_millions=(milk_produced/1000000)) %>% group_by(state) %>% arrange(milk_millions) %>% ggplot(mapping = aes(x=reorder(state,milk_millions), y=milk_millions, fill = milk_millions, color = state)) + geom_col() + coord_flip() + scale_fill_gradient( low = "red", high = "blue") +  labs(x= "State", y = "Milk produced, in millions of pounds", title = "Milk produced by state in 1997") + theme(legend.position = "none", axis.text.y = element_text(size=6))

```


### Average Milk production by state in 1997

```{r}
milk %>% filter(year == 1997) %>% mutate(milk_millions=(milk_produced/1000000)) %>% summarize (mean(milk_millions), median(milk_millions))
```
As seen from the data above, the average milk produced per state in 1997 was 3121.82 million pounds of milk and the median amount of milk produced was 1436 million pounds of milk.

### State that produced the most milk in 1997

```{r}
milk %>% filter(year == 1997) %>% mutate(milk_millions=(milk_produced/1000000)) %>% group_by(state) %>% summarise(milk_produced_millions=max(milk_millions)) %>% arrange(desc(milk_produced_millions))
```
As seen from the data above, California produced the most milk in 1997, producing 27,582 million pounds.

### State that produced the least milk in 1997

```{r}
milk %>% filter(year == 1997) %>% mutate(milk_millions=(milk_produced/1000000)) %>% group_by(state) %>% summarise(milk_produced_millions=max(milk_millions)) %>% arrange(milk_produced_millions)
```
As seen from the data above, Alaska produced the least amount of milk in 1997, producing 15 million pounds.

# Team Section 

As a team, select all of the states in which at least one of your group members has lived in. If there are less than 5, choose additional states so that you select 5 of them. For these states, plot a yearly time series for the milk production (in millions of pounds). **On this plot, display a line showing the average milk produced each year for all states.** Display this plot on the front page and comment on what information the plot conveys.

Then using the whole dataset, report the following on the front page.

- The year when the most milk was produced in the United States.
- The year when the least milk was produced in the United States.
- In 2017, report the 5 states that produced the most milk. (Hint: Use arrange)
- In 2017, report the 5 states that produced the least milk. (Hint: Use arrange)

After everyone is finished, as a team, write a few sentences in the team report in R Markdown detailing what you found to be the most helpful in learning about this data. Include all individual work in this document as well.
The final task is to reflect on the ethics of this data analysis. Brainstorm with your team on who could benefit most from this analysis? Who could suffer? Write a few sentences explaining who could benefit and suffer from this analysis and why.

## Milk production across states our group has lived in 

```{r}
states<-c("Colorado", "New Jersey","Massachusetts", "Utah", "Washington", "California","Maryland")
```

```{r}
milk_mod<-milk%>%mutate(milk_million=milk_produced/1000000)
milk_states<- milk_mod %>% filter(state %in% states)
milk_sum <- milk_mod%>% group_by(year)%>% summarise(milk_average=mean(milk_million))
ggplot()+geom_point(data=milk_states,aes(x=year,y=milk_million,color=state))+
  geom_smooth(data=milk_sum,mapping=aes(x=year,y=milk_average,color="Average all States"),se=FALSE)+
  labs(xlab="Year",y="Milk Produced (Millions of Pounds)",title="Milk Production by Year")

```

This plot shows a clear difference between California's milk production and every other state shown above, with California having by far the most milk production compared to any of the other states. This can also be seen by the average across all of the states lived in by our group, as the line would be much closer to the center of the other 6 states if not for California's massive amount of production. This plot also shows a clear upwards trend in milk production across most states as time goes on, displaying the growing demand for milk that comes with a growing population and growing US exports of milk to other countries. 


```{r}
milk_by_year<-milk_mod%>%group_by(year)%>%summarise(milk_total_million=sum(milk_million))%>%
  arrange(milk_total_million)
```

## The year in which the most milk was produced

```{r}
milk_by_year[length(milk_by_year$year),]
```

## The year in which the least milk was produced

```{r}
milk_by_year[1,]
```

## The top 5 and bottom 5 milk producers in 2017 by state

```{r}
milk2017<-milk_mod%>%filter(year==2017)
milk2017_bottom<-milk2017%>%group_by(state)%>%summarise(milk_produced=milk_million)%>%
  filter(rank(milk_produced)<=5)%>%arrange(desc(milk_produced))
milk2017_top<-milk2017%>%group_by(state)%>%summarise(milk_produced=milk_million)%>%
  filter(rank(desc(milk_produced))<=5)%>%arrange(desc(milk_produced))
milk2017_top
  
```


```{r}
milk2017_bottom
```

## Final Discussion

*Luke Fanning:* I found the in class activities relating to DIA incredibly helpful in learning how to perfomr tasks related to this data. I also found that this data allowed us to use a large amount of dplyr functions, which personally helped me gain a deeper understanding of how each one worked and allowed me to better understand how the functions connect to each other and work together. In my individual section, I separated the data down to 1997 milk production, created a new variable that represented the milk production in millions of pounds, and then used a geom_col function to display the data. I reordered the states by the variable I created, and then graphed them against each other while flipping the axes to better display the labels for each state. Finally, I added labels, reduced the font size along the y axis, and added fill and color options to better differentiate each states individual milk production
I believe that the most beneficial use of this data would be for larger milk production companies, as many can see firstly which states are most successful at producing milk (and therefore have the most productive space available for cows), and secondly which areas lack milk production of their own in order to expand markets for shipping in those regions. However, this does have a downside of hurting smaller farmers and ranchers in these lower production states, as with expanded shipping from larger companies, it will much harder for ranchers in Alaska to produce milk at all without being undercut by larger competition. 












